---
title: "Brassicas_2021_morphology"
author: "Benjamin A Taylor"
date: "20/09/2021"
output:   
  html_document:
    code_folding: hide
    self_contained: FALSE
    fig_width: 10
    fig_height: 10 
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = T, 
                      cache.path =  "/home/benjamin/Documents/Brassicas_repo/Brassicas_cache/Cache")
knitr::opts_knit$set(root.dir = "/home/benjamin/Documents/Brassicas_repo")
```

```{r load-morphology-libraries, eval = TRUE, echo = FALSE, include = FALSE}
# get libraries
basic_libraries = c("tidyverse",
                    "RColorBrewer",
                    "reshape2",
                    "DHARMa",
                    "lme4",
                    "lmerTest",
                    "grid",
                    "gridExtra")
for(lib in basic_libraries){
  if(require(package = lib, character.only = TRUE)){
    print("Successful")
  }else{print("Installing")
    install.packages(lib, Ncpus = 6)
    library(lib, character.only = TRUE)}
}
#prevent other packages overriding dplyr's select
select = dplyr::select
```

## Data import, pre-filtering and QC

```{r import-morphology-data}
#import data
phenodata.gen2.clean = read.delim("Data/Phenotypic_data/morphology_data_gen2_clean.csv", sep = ",", header = T, row.names = NULL)
#refactor to set reference levels
phenodata.gen2.clean$Wild_Dom = factor(phenodata.gen2.clean$Wild_Dom, c("Wild","Cultivated"))
phenodata.gen2.clean$Environment = factor(phenodata.gen2.clean$Environment, c("wheat","control"))

# take note of the most relevant measured variables
measure.vars = c("Days_germ",
                 "Height_1820",
                 "Height_2326",
                 "Leaf_length_1820",
                 "Leaf_length_2326",
                 "Num_leaves_1820",
                 "Num_leaves_2326",
                 "Sla_1820",
                 "Sla_2326",
                 "Sldw_1820",
                 "Sldw_2326",
                 "Aboveground_dw",
                 "Root_dw",
                 "Root_to_shoot_ratio"
) 

#narrow to spp
phenodata.gen2.clean.brass = subset(phenodata.gen2.clean, substr(Species,1,3)=="Bra")
phenodata.gen2.clean.raph = subset(phenodata.gen2.clean, substr(Species,1,3)=="Rap")
```

Check trait distributions for Brassica:

```{r brass-distribution-qc}
#Begin by checking distributions for each variable, to check nothing fishy is going on
brassmelt = melt(phenodata.gen2.clean.brass, id.vars = "Population", measure.vars = measure.vars)
gg.brass.hist = ggplot(brassmelt, aes(x=value))+
  geom_histogram() +
  facet_wrap(~variable, scales = "free")
gg.brass.hist
```

And for Raphanus:

```{r raph-distribution-qc}
#Repeat for raph
raphmelt = melt(phenodata.gen2.clean.raph, id.vars = "Population", measure.vars = measure.vars)
gg.raph.hist = ggplot(raphmelt, aes(x=value))+
  geom_histogram() +
  facet_wrap(~variable, scales = "free")
gg.raph.hist
```

#what's going on with root dry weight in these plots? It looks unimodal, but the scale seems off, perhaps indicating outliers. Sure enough,if we look at the quantile distributions for dry weight there are definitely outliers here, which we can reasonably attribute to experimental error given the sensitivity of the weighing to e.g. residual moisture in the roots. 

```{r root-dw-quantiles}
quantile(phenodata.gen2.clean.brass$Root_dw, c(seq(0,0.75,0.25),0.95,1), na.rm=T)
quantile(phenodata.gen2.clean.raph$Root_dw, c(seq(0,0.75,0.25),0.95,1), na.rm=T)
```

To be safe, we'll remove values for root dry weight that are more than 3 standard deviations above the mean.

```{r remove-dw-outliers}
#quick function to remove values 3+ SDs above the mean
crude_outliercheck = function(x){
  sd = sd(x, na.rm=T)
  mean = mean(x, na.rm=T)
  upper = (mean + sd*3)
  print("Removing values: ");print(paste(x[which(x>upper)]))
  x[which(x>upper)] = NA
  return(x)
}
#apply function to rmeove these obvious outliers
phenodata.gen2.clean.brass$Root_dw = crude_outliercheck(phenodata.gen2.clean.brass$Root_dw)
phenodata.gen2.clean.raph$Root_dw = crude_outliercheck(phenodata.gen2.clean.raph$Root_dw)
#also set root-shoot ratio to NA for these outliers
phenodata.gen2.clean.brass$Root_to_shoot_ratio[which(is.na(phenodata.gen2.clean.brass$Root_dw))] = NA
phenodata.gen2.clean.raph$Root_to_shoot_ratio[which(is.na(phenodata.gen2.clean.raph$Root_dw))] = NA
```

How do the disributions look now for Brassica and for Raphanus?

```{r distribution-qc-secondpass}
#Re-check distributions for each variable
brassmelt = melt(phenodata.gen2.clean.brass, id.vars = "Population", measure.vars = measure.vars)
gg.brass.hist = ggplot(brassmelt, aes(x=value))+
  geom_histogram() +
  facet_wrap(~variable, scales = "free")
gg.brass.hist
#Repeat for raph
raphmelt = melt(phenodata.gen2.clean.raph, id.vars = "Population", measure.vars = measure.vars)
gg.raph.hist = ggplot(raphmelt, aes(x=value))+
  geom_histogram() +
  facet_wrap(~variable, scales = "free")
gg.raph.hist
```

Root dry weight looks better now, but root-shoot ratio is still very skewed- not surprising for a ratio like this. If we log the ratio, it looks much better:

```{r plot-logged-ratios}
hist(log(phenodata.gen2.clean.brass$Root_to_shoot_ratio)); hist(log(phenodata.gen2.clean.raph$Root_to_shoot_ratio))
#okay, let's create a logged ratio
phenodata.gen2.clean.brass$Log_root_shoot_ratio = log(phenodata.gen2.clean.brass$Root_to_shoot_ratio)
phenodata.gen2.clean.raph$Log_root_shoot_ratio = log(phenodata.gen2.clean.raph$Root_to_shoot_ratio)
#update the measure.vars object (also drop traits that remain clearly unimodal)
measure.vars.brass = c(measure.vars[!(measure.vars %in% c("Days_germ","Root_to_shoot_ratio"))],"Log_root_shoot_ratio")
measure.vars.raph = c(measure.vars[!(measure.vars %in% c("Days_germ","Root_to_shoot_ratio"))],"Log_root_shoot_ratio")

```

We'll use logged root-shoot ratio instead of the unmodified ratio henceforth. We'll also drop days to germination, since this is clearly a highly unimodal trait. 

## Trait model testing

We'd now like to fit the data for each trait so that we can plug our data into mixed models down the line. In an ideal world each trait would be fit reasonably well by a normal distribution- let's see if that's the case. What we're looking for: 

1. Points fitted nicely along the QQ plot abline
2. Even distribution about abline in fitted vs residuals plot
3. Roughly horizontal fit for scale-location plot
4. No outiers in residuals vs leverage plot

First for Brassica traits:

```{r modeltesting-gauss-brassica, fig.show="hold", results='asis'}
for(i in 1:length(measure.vars.brass)){
  
  response.var = measure.vars.brass[i]
  phenodata.gen2.clean.brass.omit = subset(phenodata.gen2.clean.brass, is.na(get(response.var))==F)
  model.gauss = lmer(scale(get(response.var))~Environment + Wild_Dom + Environment:Wild_Dom + (1|Population), 
                     data=phenodata.gen2.clean.brass.omit, 
                     control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))

  #qq plot
  plot.qq = qqmath(model.gauss,
                   main="Q-Q plot", font.main=1)
  #fitted vs residuals plot
  plot.fitres = plot(model.gauss, type=c("p","smooth"), col.line=1, 
                     main="Fitted vs residuals", font.main=1)
  #scale-location plot
  plot.scaloc = plot(model.gauss,
     sqrt(abs(resid(.)))~fitted(.),
     type=c("p","smooth"), col.line=1, 
     main="Scale-location",font.main=1)
  #residuals vs leverage
  plot.reslev = plot(model.gauss, rstudent(.) ~ hatvalues(.),
                     main="Residuals vs leverage",font.main=1)
  #return plots
  grid.arrange(plot.qq,
               plot.fitres,
               plot.scaloc,
               plot.reslev,
               top=grid::textGrob(response.var, gp=grid::gpar(fontsize=18,font=2)))

}

```

Then for Raphanus traits:

```{r modeltesting-gauss-raphanus, fig.show="hold", results='asis'}
for(i in 1:length(measure.vars.raph)){
  
  response.var = measure.vars.raph[i]
  phenodata.gen2.clean.raph.omit = subset(phenodata.gen2.clean.raph, is.na(get(response.var))==F)
  model.gauss = lmer(scale(get(response.var))~Environment + Wild_Dom + Environment:Wild_Dom + (1|Population), 
                     data=phenodata.gen2.clean.raph.omit, 
                     control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))

  #qq plot
  plot.qq = qqmath(model.gauss,
                   main="Q-Q plot", font.main=1)
  #fitted vs residuals plot
  plot.fitres = plot(model.gauss, type=c("p","smooth"), col.line=1, 
                     main="Fitted vs residuals", font.main=1)
  #scale-location plot
  plot.scaloc = plot(model.gauss,
     sqrt(abs(resid(.)))~fitted(.),
     type=c("p","smooth"), col.line=1, 
     main="Scale-location",font.main=1)
  #residuals vs leverage
  plot.reslev = plot(model.gauss, rstudent(.) ~ hatvalues(.),
                     main="Residuals vs leverage",font.main=1)
  #return plots
  grid.arrange(plot.qq,
               plot.fitres,
               plot.scaloc,
               plot.reslev,
               top=grid::textGrob(response.var, gp=grid::gpar(fontsize=18,font=2)))

}

```


While none of these fits are perfect, the only massive issues are with number of leaves, since this is a discrete integer trait and thus fits weirdly to a gaussian distribution. For this trait, we might be better off fitting a Poisson distribution:

```{r modeltesting-poisson-brassica, fig.show="hold", results='asis', fig.height = 5}
discrete.traits = c("Num_leaves_1820","Num_leaves_2326")

for(i in 1:length(discrete.traits)){
  
  response.var = discrete.traits[i]
  phenodata.gen2.clean.brass.omit = subset(phenodata.gen2.clean.brass, is.na(get(response.var))==F)
  model.gauss = glmer(get(response.var)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Population), 
                     data=phenodata.gen2.clean.brass.omit, 
                     control = glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)),family = poisson())

  #qq plot
  plot.qq = qqmath(model.gauss,
                   main="Q-Q plot", font.main=1)
  #fitted vs residuals plot
  plot.fitres = plot(model.gauss, type=c("p","smooth"), col.line=1, 
                     main="Fitted vs residuals", font.main=1)
  #return plots
  grid.arrange(plot.qq,
               plot.fitres,
               ncol=2,
               top=grid::textGrob(response.var, gp=grid::gpar(fontsize=18,font=2)))

}

```

This doesn't necessarily look better, but it does make more sense (number of leaves is much more likely to be poisson distributed than gaussian, since it's a discrete trait) so for num leaves we'll use a poisson distribution going forward in place of a gaussian. 

The other trait that looks odd in the above diagnostic plots is root dry-weight. We already know that this trait suffers from some skew, but we'd hoped to improve that by removing outliers earlier. Perhaps some kind of inverse link function could help with this?

```{r modeltesting-inverse-brassica, fig.show="hold", results='asis', fig.height = 5}
  response.var = "Root_dw"
  phenodata.gen2.clean.brass.omit = subset(phenodata.gen2.clean.brass, is.na(get(response.var))==F)
  model.gauss = glmer(get(response.var)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Population), 
                     data=phenodata.gen2.clean.brass.omit, 
                     control = glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)),family=gaussian(link="inverse"))
  #qq plot
  plot.qq = qqmath(model.gauss,
                   main="Q-Q plot", font.main=1)
  #fitted vs residuals plot
  plot.fitres = plot(model.gauss, type=c("p","smooth"), col.line=1, 
                     main="Fitted vs residuals", font.main=1)
  #return plots
  grid.arrange(plot.qq,
               plot.fitres,
               ncol=2,
               top=grid::textGrob(response.var, gp=grid::gpar(fontsize=18,font=2)))
```

That doesn't look any better. What about a gamma distribution?

```{r modeltesting-gamma-brassica, fig.show="hold", results='asis', fig.height = 5}
model.gauss = glmer(get(response.var)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Population), 
                     data=phenodata.gen2.clean.brass.omit, 
                     control = glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)),family=Gamma())

  #qq plot
  plot.qq = qqmath(model.gauss,
                   main="Q-Q plot", font.main=1)
  #fitted vs residuals plot
  plot.fitres = plot(model.gauss, type=c("p","smooth"), col.line=1, 
                     main="Fitted vs residuals", font.main=1)
  #return plots
  grid.arrange(plot.qq,
               plot.fitres,
               ncol=2,
               top=grid::textGrob(response.var, gp=grid::gpar(fontsize=18,font=2)))
```

Even worse! Okay, there doesn't seem to be any way to fit this trait nicely, so let's just model it with a gaussian distribution but take the results with a pinch of salt.
