---
title: "Brassicas_2021_summarized_results"
author: "Benjamin A Taylor"
date: "25/09/2021"
output:   
  html_document:
    code_folding: hide
    self_contained: FALSE
    fig_width: 10
    fig_height: 10 
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = F, 
                      cache.path =  "/home/benjamin/Documents/Brassicas_repo/Brassicas_cache/Cache")
knitr::opts_knit$set(root.dir = "/home/benjamin/Documents/Brassicas_repo")
```

```{r load-morphology-libraries, eval = TRUE, echo = FALSE, include = FALSE}
# get libraries
basic_libraries = c("tidyverse",
                    "RColorBrewer",
                    "grid",
                    "gridExtra",
                    "lattice")
for(lib in basic_libraries){
  if(require(package = lib, character.only = TRUE)){
    print("Successful")
  }else{print("Installing")
    install.packages(lib, Ncpus = 6)
    library(lib, character.only = TRUE)}
}
#prevent other packages overriding dplyr's select
select = dplyr::select
#set working directory
setwd("/home/benjamin/Documents/Brassicas_repo")
```

This markdown document summarizes the principal results of the Brassica project generation 2 analyses. Analyses are not performed within this document; prior to running this file, the results should be generated by running the two other .Rmd files in this repository ("Brassicas_2021_morphology.Rmd" and "Brassicas_2021_rnaseq.Rmd"). 

## Work package 1: Comparing domesticates to their wild progenitors

For this work package, we wish to compare wild progenitor species to modern domesticates. For Brassica, the comparison is between domesticated B rapa with wild B rapa, while for Raphanus the comparison is between R raphanistrum (wild) and all other Raphanus samples, all of which appear to be domesticated.

# Brassica gene expression

For the gene expression analysis, out model design is:

  $$gene expression \sim domestication + treatment + domestication*treatment$$
Results summary (number of genes and GO terms enriched for each dependent variable) for wild vs domesticated Brassica rapa:

```{r}
read.csv("Analysis/RNAseq/Tables/brassica_deseq_domesticate_vs_progenitor_summary.csv", row.names = 1)
```

We can see from this that a great many genes are associated with domestication, but few with stress. 41 genes show an interaction between treatment and domestication, and are therefore of particular interest. Summary of the 25 GO terms associated with these interaction genes: