---
title: "Brassicas_2021"
author: "Benjamin A Taylor"
date: "29/03/2021"
output:   
  html_document:
    code_folding: hide
    self_contained: FALSE
    fig_width: 10
    fig_height: 10 
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = T, cache.path =  "/home/benjamin/Documents/Brassicas_repo/Brassicas_cache/Cache")
```
```{r load-libraries, eval = TRUE, echo = FALSE, include = FALSE}
# get libraries
basic_libraries <- c("tidyverse",
                     "RColorBrewer",
                     "lme4",
                     "lmerTest",
                     "performance",
                     "see",
                     "factoextra",
                     "reshape2",
                     "ggpubr")

for (lib in basic_libraries) {
        if (require(package = lib, character.only = TRUE)) {
                print("Successful")
        } else {
                print("Installing")
                install.packages(lib, Ncpus = 6)
                library(lib, character.only = TRUE )
        }
}

#prevent other packages overriding dplyr's select
select = dplyr::select
```

Import morphology data and clean up the formatting:

```{r import-data, eval = TRUE, echo = FALSE, include = FALSE}
options(stringsAsFactors = F)
setwd("/home/benjamin/Documents/Brassicas_repo")

# import data
phenodata.gen2 = read.delim("Data/Phenotypic_data/morphology_data_gen2.txt", sep = ",", header = T, row.names = NULL)
colnames(phenodata.gen2)
table(phenodata.gen2$Species)

# remove unused columns
phenodata.gen2 = select(phenodata.gen2, -c("Total_la"))

# change the environment labels to something less verbose
phenodata.gen2$Environment =  ifelse(phenodata.gen2$Environment == "wheat competition", "wheat", "control")
# for now simplify down to wild vs domesticated
phenodata.gen2$Wild_Dom =  ifelse(phenodata.gen2$Wild_Dom == "Wild", "Wild", "Cultivated")

# for some factors, we need to take into account timepoint of measurement
table(phenodata.gen2$Days_meas_taken)
phenodata.gen2$DayMeasured = ifelse(phenodata.gen2$Days_meas_taken %in% c(18:20,23:26), 
                                    ifelse(phenodata.gen2$Days_meas_taken %in% c(18:20), 
                                           "d18to20", "d23to26"), NA)
#subset height, num leaves, sla, sldw, leaf length based on measure data
repeated.measures = c("Height","Leaf_length","Num_leaves","Sla","Sldw")
#new cols for days 18-20
newcols = select(phenodata.gen2, repeated.measures)
newcols[which(phenodata.gen2$DayMeasured != "d18to20"),] = NA
colnames(newcols) = paste0(colnames(newcols),"_1820")
phenodata.gen2 = cbind(phenodata.gen2, newcols)
#new cols for days 23-26
newcols = select(phenodata.gen2, repeated.measures)
newcols[which(phenodata.gen2$DayMeasured != "d23to26"),] = NA
colnames(newcols) = paste0(colnames(newcols),"_2326")
phenodata.gen2 = cbind(phenodata.gen2, newcols)
#remove old, now-defunct measure cols
phenodata.gen2 = select(phenodata.gen2, -repeated.measures)

# take note of the most relevant measured variables
measure.vars = c("Days_germ", #days from potting to germination
                 "Height_1820", #max distance from soil to plant
                 "Height_2326",
                 "Leaf_length_1820", #base to tip of longest non-cotyledon
                 "Leaf_length_2326",
                 "Num_leaves_1820", #number of leaves
                 "Num_leaves_2326", 
                 "Sla_1820", #single leaf area of oldest non-cotyledon
                 "Sla_2326",
                 "Sldw_1820", #single lead dry weight of oldest non-cotyledon
                 "Sldw_2326",
                 #"Sp_la", #specific leaf area
                 "Aboveground_dw", #all biomass above soil
                 #"Day_flowering", #days from germination to flowering
                 "Root_dw", #all biomass below soil
                 #"Total_biomass", #sum of above and below ground biomass
                 "Root_to_shoot_ratio" #ratio of below to above-ground biomass (I think)
) 

# convert to numeric where applicable
phenodata.gen2.clean =  mutate_at(phenodata.gen2, measure.vars, function(x) as.numeric(as.character(x))) %>%
  select(c("Label_front","Species","Population","Wild_Dom","Environment",measure.vars)) 



#refactor to set reference levels
phenodata.gen2.clean$Wild_dom <- factor(phenodata.gen2.clean$Wild_Dom,
                                        c("Wild","Cultivated"))
```

Remove extreme outlier measurements:

```{r}
# looks like we have a pretty extreme outlier for root dry weight
summary(phenodata.gen2.clean$Root_dw)
ggqqplot(phenodata.gen2.clean$Root_dw)
hist(phenodata.gen2.clean$Root_dw,
  breaks = sqrt(length(phenodata.gen2.clean$Root_dw)))
#remove the outlier
phenodata.gen2.clean$Root_dw[which(phenodata.gen2.clean$Root_dw == max(phenodata.gen2.clean$Root_dw,na.rm = T))] = NA


# per PCA, the below sample is a heavy outlier for Raphanus
phenodata.gen2.clean = phenodata.gen2.clean[-c(which(phenodata.gen2.clean$Label_front == "2.0-RRA-WP2-C-849")),]
```


Wel'll begin with a simple exploration of the  data. The proportion of the samples for which we actually have complete data is actually quite small:

```{r}
phenodata.gen2.clean.brass = phenodata.gen2.clean %>% subset(substr(Species,1,8)=="Brassica")
print(paste0(length(which(complete.cases(phenodata.gen2.clean.brass))),"/",nrow(phenodata.gen2.clean.brass)," of Brassica records are complete"))

phenodata.gen2.clean.raph = phenodata.gen2.clean %>% subset(substr(Species,1,8)=="Raphanus")
print(paste0(length(which(complete.cases(phenodata.gen2.clean.raph))),"/",nrow(phenodata.gen2.clean.raph)," of Raphanus records are complete"))
```
Nonetheless, this sample will hopefully be representative for an initial exploration. Looking at the analyses that Jasmine has generated so far, two things are notable: first, we have a large number of measured morphological characteristics, which means that any stats we generate are hard to interpret and likely to be diminished significantly by FDR correction; and second, the results generated for individual characteristics are inconsistent and marginal. Also, individual traits are likely highly correlated, which again means we might not want to regress against them individually:

```{r}
phenodata.gen2.clean.compcases = phenodata.gen2.clean[complete.cases(phenodata.gen2.clean),]

brassica.cors = phenodata.gen2.clean.compcases %>% 
  subset(substr(Species,1,8)=="Brassica") %>%
  select(measure.vars) %>%
  cor() %>% round(2)

brassica.cors[upper.tri(brassica.cors)] = NA

brassica.cors.plot = ggplot(data = melt(brassica.cors, na.rm = T), aes(x=Var1, y=Var2, fill=value)) + 
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1),
       axis.text.y = element_text(size = 12),
       axis.title = element_blank())+
 coord_fixed()

raphanus.cors = phenodata.gen2.clean.compcases %>% 
  subset(substr(Species,1,8)=="Raphanus") %>%
  select(measure.vars) %>%
  cor() %>% round(2)

raphanus.cors[upper.tri(raphanus.cors)] = NA

raphanus.cors.plot = ggplot(data = melt(raphanus.cors, na.rm = T), aes(x=Var1, y=Var2, fill=value)) + 
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1),
       axis.text.y = element_text(size = 12),
       axis.title = element_blank())+
 coord_fixed()

cors.arrange = ggarrange(brassica.cors.plot, raphanus.cors.plot, 
                         common.legend = T,labels = c("Brassica","Raphanus"))

ggsave(cors.arrange, filename = "phenotypic_correlations.pdf",
       device = "pdf", path = "Analysis/Phenotypic_analysis/Images",
       width =  18, height = 11, units = "cm")
```

```{r}
phenodata.gen2.clean.brass = subset(phenodata.gen2.clean, substr(Species,1,8)=="Brassica")

# phenodata.gen2.clean.brass$Days_germ
# 
# response_var = measure.vars[1]
# response_var
# foo = lmer(scale(get(response_var))~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
#              data=phenodata.gen2.clean.brass, 
#              control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
#check_model(foo)

#loop through measure vars and regress each one against treatment vars
for(i in 1:length(measure.vars)){
  
  #get response of interest
  response_var = measure.vars[i]
  #regress against environment and treatment and extract p-vals
  # foo = lmer(scale(get(response_var))~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
  #            data=phenodata.gen2.clean.brass, 
  #            control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
  foo = glmer(get(response_var)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population),
      data=phenodata.gen2.clean.brass,family="Gamma",glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
  #check_model(foo)
  foo = summary(foo)$coefficients 
  #if first loop, instantiate df to store outputs
  if(i==1){
    brassica.coefs = data.frame(row.names = row.names(foo))
    brassica.vals = data.frame(row.names = row.names(foo))
    }
  #cbind coefs to df
  #brassica.coefs = cbind(brassica.coefs,foo[,5])
  brassica.coefs = cbind(brassica.coefs,foo[,4])
  brassica.vals = cbind(brassica.vals,foo[,1])

  #if last loop, give colnames
  if(i==length(measure.vars)){colnames(brassica.coefs)=measure.vars}
  

}

#get labels for plot
brassica.coef.labels = paste0(as.matrix(signif(brassica.coefs,2)),"\n(",as.matrix(signif(brassica.vals,2)),")")
#coerce df for ggplot and apply FDR correction
brassica.coefs.fdr = reshape2::melt(rownames_to_column(brassica.coefs)) %>% mutate(value = p.adjust(value, method = "BH"))

#plot
brassica.traits.fdr.plot = ggplot(data = brassica.coefs.fdr, aes(x = variable, y = rowname, fill = value)) +
  geom_tile(color = "grey50", size = 0.5) +
  #geom_text(aes(label=round(value,4)), size = 3, face = "bold") +
  geom_tile(aes(fill = value),color = "gray", size=.75, width=1, height = 1) +
  geom_text(aes(label=c(brassica.coef.labels), 
            lineheight = 0.75, size = 2)) +
  scale_fill_gradientn(colours = colorRampPalette(rev(c("#FFFFFF",brewer.pal(n = 9, name = "Reds")[1:5])),bias=6)(20),
                       breaks = c(0.0,0.05,1),
                       expand = c(0,0),
                       limits = c(0,1),
                       guide = guide_colourbar(barheight = 25,
                                               #title = "p-value\n(adjusted)",
                                               title = "p-value\n(adjusted)",
                                               title.vjust = 2,
                                               frame.colour = "black", 
                                               frame.linewidth = 1.5)) +
  theme_minimal() + 
  theme(#aspect.ratio = 1,
        panel.grid = element_line(size = 0.2, colour = "gray80"),
        axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1),
        axis.text.y = element_text(size = 12),
        axis.title = element_blank()#,
        #legend.text = element_text(size=12),
        #legend.title = element_text(size=15, face = "bold")
        ) +
  coord_fixed()

```

```{r}
phenodata.gen2.clean.raph = subset(phenodata.gen2.clean, substr(Species,1,8)=="Raphanus")

phenodata.gen2.clean.raph$Days_germ

response_var = measure.vars[1]
response_var
foo = lmer(scale(get(response_var))~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
             data=phenodata.gen2.clean.raph, 
             control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
#check_model(foo)

#loop through measure vars and regress each one against treatment vars
for(i in 1:length(measure.vars)){
  
  #get response of interest
  response_var = measure.vars[i]
  #regress against environment and treatment and extract p-vals
  # foo = lmer(scale(get(response_var))~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
  #            data=phenodata.gen2.clean.raph, 
  #            control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
  foo = glmer(get(response_var)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population),
      data=phenodata.gen2.clean.raph,family="Gamma",glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
  #check_model(foo)
  foo = summary(foo)$coefficients 
  #if first loop, instantiate df to store outputs
  if(i==1){
    raphanus.coefs = data.frame(row.names = row.names(foo))
    raphanus.vals = data.frame(row.names = row.names(foo))
    }
  #cbind coefs to df
  #raphanus.coefs = cbind(raphanus.coefs,foo[,5])
  raphanus.coefs = cbind(raphanus.coefs,foo[,4])
  raphanus.vals = cbind(raphanus.vals,foo[,1])

  #if last loop, give colnames
  if(i==length(measure.vars)){colnames(raphanus.coefs)=measure.vars}
  

}

#get labels for plot
raphanus.coef.labels = paste0(as.matrix(signif(raphanus.coefs,2)),"\n(",as.matrix(signif(raphanus.vals,2)),")")
#coerce df for ggplot and apply FDR correction
raphanus.coefs.fdr = reshape2::melt(rownames_to_column(raphanus.coefs)) %>% mutate(value = p.adjust(value, method = "BH"))

#plot
raphanus.traits.fdr.plot = ggplot(data = raphanus.coefs.fdr, aes(x = variable, y = rowname, fill = value)) +
  geom_tile(color = "grey50", size = 0.5) +
  #geom_text(aes(label=round(value,4)), size = 3, face = "bold") +
  geom_tile(aes(fill = value),color = "gray", size=.75, width=1, height = 1) +
  geom_text(aes(label=c(raphanus.coef.labels), 
            lineheight = 0.75, size = 2)) +
  scale_fill_gradientn(colours = colorRampPalette(rev(c("#FFFFFF",brewer.pal(n = 9, name = "Reds")[1:5])),bias=6)(20),
                       breaks = c(0.0,0.05,1),
                       expand = c(0,0),
                       limits = c(0,1),
                       guide = guide_colourbar(barheight = 25,
                                               #title = "p-value\n(adjusted)",
                                               title = "p-value\n(adjusted)",
                                               title.vjust = 2,
                                               frame.colour = "black", 
                                               frame.linewidth = 1.5)) +
  # scale_x_discrete(labels = c("Domestication",
  #                             "Dom.(European)",
  #                             "Domestication:Environment",
  #                             "Domestication:Environment:Origin",
  #                             "Environment",
  #                             "Intercept")) +
  theme_minimal() + 
  theme(#aspect.ratio = 1,
        panel.grid = element_line(size = 0.2, colour = "gray80"),
        axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1),
        axis.text.y = element_text(size = 12),
        axis.title = element_blank()#,
        #legend.text = element_text(size=12),
        #legend.title = element_text(size=15, face = "bold")
        ) +
  coord_fixed()
```

```{r}
traits.arrange = ggarrange(brassica.traits.fdr.plot, raphanus.traits.fdr.plot, nrow = 2, ncol = 1,
                           common.legend = F,labels = c("Brassica","Raphanus"), legend = "none")

ggsave(traits.arrange, filename = "phenotypic_glmer_pvals.pdf",
       device = "pdf", path = "Analysis/Phenotypic_analysis/Images",
       width =  25, height = 25, units = "cm")
```



My initial instinct is to look for a way to reduce the dimensionality of the data, so let's try generating PCAs. We'll start with the Brassicas.


We'll begin with a scree plot, which shows that PCs 1 and 2 together explain about two thirds of the data. There's a big drop off in information per component for PCs 3+ so we'll focus on just the first two PCs for now. 

```{r}
##pca for brassicas
# subset to brassicas and complete cases
pca.counts = phenodata.gen2.clean.compcases %>% 
  subset(substr(Species,1,8)=="Brassica") %>%
  remove_rownames() %>%
  column_to_rownames("Label_front") %>%
  select(measure.vars) #%>% 
  #scale() 

#generate PCA object
#NB: pre-scaling is important for PCA, in order to prevent high-variance covariates dominating the PC calculations
data.pca = prcomp(pca.counts,scale. = T)

#generate scree plot
fviz_eig(data.pca)
```
```{r}
##pca for raphanus
# subset to brassicas and complete cases
pca.counts.raph = phenodata.gen2.clean.compcases %>% 
  subset(substr(Species,1,8)=="Raphanus") %>%
  remove_rownames() %>%
  column_to_rownames("Label_front") %>%
  select(measure.vars) #%>% 
  #scale() 

#generate PCA object
#NB: pre-scaling is important for PCA, in order to prevent high-variance covariates dominating the PC calculations
data.pca.raph = prcomp(pca.counts.raph,scale. = T)

#generate scree plot
fviz_eig(data.pca.raph)
```


Looking at the loading of these two PCs, we can see that they decompose fairly cleanly into a component that represents leaf size/length/weight, and a second component that represents other metrics related to size and weight. Notably, individuals that germinated more quickly are also larger. We also predict that time to flowering will be loaded in the opposite direction to size, but because we have very few complete cases when including flowering time, those data aren't in this plot. 

```{r}
#visualize PC loadings
pca.var.plot = fviz_pca_var(data.pca,
                            col.var = "contrib", # Color by contributions to the PC
                            gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                            repel = TRUE,     # Avoid text overlapping
                            title = "Brassica")

#visualize PC loadings
pca.var.plot.raph = fviz_pca_var(data.pca.raph,
                                 col.var = "contrib", # Color by contributions to the PC
                                 gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                                 repel = TRUE ,    # Avoid text overlapping
                                 title = "Raphanus")

pca.var.arrange = ggarrange(pca.var.plot, pca.var.plot.raph, common.legend = T)

ggsave(pca.var.arrange, filename = "PCA_loadings.pdf",
       device = "pdf", path = "Analysis/Phenotypic_analysis/Images",
       width =  25, height = 15, units = "cm")
```

Now let's plot individuals on these PCs and color according either to species or to wild vs domesticated origins:

```{r}
#plot by species
fviz_pca_ind(data.pca,
             habillage = subset(phenodata.gen2.clean.compcases, substr(Species,1,8)=="Brassica")$Species, #label by spp
             label = F, 
             pointsize = 3.5,
             invisible = "quali", #removes centre point for groups from plot
             ggtheme = theme_gray()) + #grey background makes picking out patterns easier
  scale_shape_manual(values = rep(20,7)) #hacky way of preventing the plot from using different shapes for each group

#plot by environment of origin
pca.ind.brass.wildDom = fviz_pca_ind(data.pca,
             habillage = subset(phenodata.gen2.clean.compcases, substr(Species,1,8)=="Brassica")$Wild_Dom, #label by spp
             label = F, 
             pointsize = 3.5,
             title = "\t", #hacky way of removing plot title
             invisible = "quali", #removes centre point for groups from plot
             ggtheme = theme_gray()) + #grey background makes picking out patterns easier
  scale_shape_manual(values = rep(20,7)) #hacky way of preventing the plot from using different shapes for each group

#plot by treatment
pca.ind.brass.treatment = fviz_pca_ind(data.pca,
             habillage = subset(phenodata.gen2.clean.compcases, substr(Species,1,8)=="Brassica")$Environment, #label by spp
             label = F, 
             col.ind = c("blue","green"),
             pointsize = 3.5,
             title = "\t", #hacky way of removing plot title
             invisible = "quali", #removes centre point for groups from plot
             ggtheme = theme_gray()) + #grey background makes picking out patterns easier
  scale_shape_manual(values = rep(20,7)) + #hacky way of preventing the plot from using different shapes for each group
  scale_color_brewer(palette="Dark2")

pca.ind.brass.arrange = ggarrange(pca.ind.brass.wildDom,pca.ind.brass.treatment)  

ggsave(pca.ind.brass.arrange, filename = "PCA_scatters_brass.pdf",
       device = "pdf", path = "Analysis/Phenotypic_analysis/Images",
       width =  25, height = 15, units = "cm")
```

```{r}
#plot by species
fviz_pca_ind(data.pca.raph,
             habillage = subset(phenodata.gen2.clean.compcases, substr(Species,1,8)=="Raphanus")$Species, #label by spp
             label = F, 
             pointsize = 3.5,
             invisible = "quali", #removes centre point for groups from plot
             ggtheme = theme_gray()) + #grey background makes picking out patterns easier
  scale_shape_manual(values = rep(20,7)) #hacky way of preventing the plot from using different shapes for each group

#plot by environment of origin
pca.ind.raph.wildDom = fviz_pca_ind(data.pca.raph,
             habillage = subset(phenodata.gen2.clean.compcases, substr(Species,1,8)=="Raphanus")$Wild_Dom, #label by spp
             label = F, 
             pointsize = 3.5,
             title = "\t", #hacky way of removing plot title
             invisible = "quali", #removes centre point for groups from plot
             ggtheme = theme_gray()) + #grey background makes picking out patterns easier
  scale_shape_manual(values = rep(20,7)) #hacky way of preventing the plot from using different shapes for each group

#plot by treatment
pca.ind.raph.treatment = fviz_pca_ind(data.pca.raph,
             habillage = subset(phenodata.gen2.clean.compcases, substr(Species,1,8)=="Raphanus")$Environment, #label by spp
             label = F, 
             pointsize = 3.5,
             title = "\t", #hacky way of removing plot title
             invisible = "quali", #removes centre point for groups from plot
             ggtheme = theme_gray()) + #grey background makes picking out patterns easier
  scale_shape_manual(values = rep(20,7))  + #hacky way of preventing the plot from using different shapes for each group
  scale_color_brewer(palette="Dark2")

pca.ind.raph.arrange = ggarrange(pca.ind.raph.wildDom,pca.ind.raph.treatment)  

ggsave(pca.ind.raph.arrange, filename = "PCA_scatters_raph.pdf",
       device = "pdf", path = "Analysis/Phenotypic_analysis/Images",
       width =  25, height = 15, units = "cm")
```

We now want to ask whether individuals' assignments along the first two PCs are predicted well by cultivated stats, treatment, or their interaction. Let's begin by checking distributions of the PCs. First with PC1:

```{r}
#append values for first two PCs to pheno data
pca.out = phenodata.gen2.clean.compcases %>% 
  subset(substr(Species,1,8)=="Brassica") %>%
  mutate(pca.counts, PC1 = data.pca$x[,"PC1"], PC2 = data.pca$x[,"PC2"])

#generate an lmm for PC1
PC1.model.gaus = lmer(PC1~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), data=pca.out)

#initial model throws an error so we add a control parameter and also scale the response variable
PC1.model.gaus = lmer(scale(PC1)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
                      data=pca.out, 
                      control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
#that helped! Let's check out the model performance- it looks decent
#check_model(PC1.model.gaus)

#looking at the output, however, we have no meaningful effects:
summary(PC1.model.gaus)$coefficients

write.csv(signif(summary(PC1.model.gaus)$coefficients,3),
          file = "Analysis/Phenotypic_analysis/Tables/brass_PC1_coefs.csv")
```
Now with PC2:

```{r}
PC2.model.gaus = lmer(scale(PC2)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
                      data=pca.out, 
                      control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
#model performance is much worse for this PC
#check_model(PC2.model.gaus)
summary(PC2.model.gaus)$coefficients

write.csv(signif(summary(PC2.model.gaus)$coefficients,3),
          file = "Analysis/Phenotypic_analysis/Tables/brass_PC2_coefs.csv")

#let's try a log link to see if that improves the model at all
PC2.model.gaus.log = glmer(scale(PC2)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
                      data=pca.out, 
                      family=gaussian(link = "log"), 
                      mustart=pmax(scale(pca.out$PC2),1e-3),
                      control = glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
#this sort of works works, but the message 'boundary (singular) fit' indicates either some multicollinearity or alternatively an effect with an estimate close to zero
#also the model fit's horrible
#check_model(PC2.model.gaus.log)
summary(PC2.model.gaus.log)$coefficients
```

Now with Raph:

```{r}
#append values for first two PCs to pheno data
pca.out.raph = phenodata.gen2.clean.compcases %>% 
  subset(substr(Species,1,8)=="Raphanus") %>%
  mutate(pca.counts.raph, PC1 = data.pca.raph$x[,"PC1"], PC2 = data.pca.raph$x[,"PC2"])

#generate an lmm for PC1
PC1.raph.model.gaus = lmer(PC1~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), data=pca.out.raph)

# #initial model throws an error so we add a control parameter and also scale the response variable
# PC1.raph.model.gaus = lmer(scale(PC1)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
#                       data=pca.out.raph, 
#                       control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
# #that helped! Let's check out the model performance- it looks decent
# #check_model(PC1.model.gaus)

#looking at the output, however, we have no meaningful effects:
summary(PC1.raph.model.gaus)$coefficients

write.csv(signif(summary(PC1.raph.model.gaus)$coefficients,3),
          file = "Analysis/Phenotypic_analysis/Tables/raph_PC1_coefs.csv")
```

```{r}
#generate an lmm for PC2
PC2.raph.model.gaus = lmer(PC2~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), data=pca.out.raph)

#initial model throws an error so we add a control parameter and also scale the response variable
PC2.raph.model.gaus = lmer(scale(PC2)~Environment + Wild_Dom + Environment:Wild_Dom + (1|Wild_Dom/Population), 
                      data=pca.out.raph, 
                      control = lmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)))
#that helped! Let's check out the model performance- it looks decent
#check_model(PC2.model.gaus)

#looking at the output, however, we have no meaningful effects:
summary(PC2.raph.model.gaus)$coefficients

write.csv(signif(summary(PC2.raph.model.gaus)$coefficients,3),
          file = "Analysis/Phenotypic_analysis/Tables/raph_PC2_coefs.csv")
```


Okay, all the above is a bit moot anyway because of our preponderance of incomplete cases. By the protocol, biomass measurements were taken separately from height/leaf measurements, so it's possible we'll be able to include more data is we trat these things separately.

```{r}
phenodata.gen2.clean %>% complete.cases() %>% table()

phenodata.gen2.clean %>% select(-c("Aboveground_dw","Root_dw")) %>% complete.cases() %>% table

phenodata.gen2.clean %>% select(-c("Height","Leaf_length","Num_leaves","Sla","Sldw")) %>% complete.cases() %>% table
```

